---
title: "Factors affecting MPG in cars"
author: "Dominic Tabeta"
date: "May 9, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Executive Summary

In this report we try to answer the question "Is an automatic or manual transmission better for MPG?" and we also try to quantify the MPG difference between automatic and manual transmissions. We compare MPG of 32 cars with the following features and try to quantify the MPG impact of some of these selected features:

* Transmission (0=automatic, 1= manual)
* Number of cylinders
* Displacement (cubic inches)
* Gross horsepower
* Rear axle ratio
* weight (in 1000 lbs)
* 1/4 mile time
* Engine configuration (0=V, 1=straight)
* Number of forward gear
* Number of forward gears
* Number of carburetors

## Exploratory analysis

First we'll try to address the question "Is an automatic or manual transmission better for MPG?". We see from the below graphs that at manual cars (am=0) have higher mean and median mpg than automatic transmission cars. I have added to the below boxplots the mean mpg for each group indicated by a red point and text. At first glance the boxplot indicates that manual cars have higher mpg, but there is significant overlap of mpg between the two groups which suggest that determing the mpg of a car depends on more than just whether the transmission is manual or automatic as the am feature does not explain all the variance of mpg.
```{r am_only}
means <- aggregate(mpg ~  am, mtcars, mean)
g<-ggplot(data=mtcars,aes(y=mpg,x=factor(am)))+
        geom_boxplot()+stat_summary(fun.y=mean, colour="darkred", geom="point", shape=18, size=3,show_guide = FALSE)+geom_text(data = means, aes(label = mpg, y = mpg + 1))
g
fitamonly<-lm(mpg~am,mtcars)
summary(fitamonly)
```
A linear regression model using only am as a predictor has an R-squared value of only 0.3385, which suggests a more accurate model predicting a car's mpg should use other predictors as well.

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

pairs(mpg~.,mtcars)
#g<-ggplot(data=mtcars,aes(y=mpg,x=wt,colour=factor(am)))+geom_point()+geom_abline()
#g

#g<-ggplot(data=mtcars,aes(y=mpg,x=factor(am)))+geom_boxplot()
#g


```


